---
title: "Report"
author: "Liucheng Shi, Zhuohui Liang, Ruwen Zhou, Jiying Han"
date: "3/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)  
library(cluster)    
library(factoextra) 
library(dendextend) 
```

## Introduction

## Data Preparation

## PCA

## Apply EM algorithm

## Gene-expression signatures

## Analysing Gaussian-Mixture model with Principal Component

## Result

## Hierarchical clustering

Hierarchical clustering is an alternative approach to k-means clustering for identifying groups in the dataset. In our project, our purpose is to classify scRNA-seq into different clusters based on their gene expressions and identify potential existence of cell subtypes. We choose agglomerative hierarchical clustering method, which is a bottom-up method. Each object is initially considered as a single-element cluster. At each step of the algorithm, the two clusters that are the most similar are combined into a new bigger cluster. The iteration will not stop until all elements are being classified into one single cluster. This makes the result a tree and can be visualized as a dendrogram.

When measuring the dissimilarity between each pair of observations distance, Euclidean distance has been used. However, when measuring the dissimilarity between two clusters of observations, we have applied different methods.

* Minimum or single linkage clustering: It computes all pairwise dissimilarities between the elements in cluster 1 and the elements in cluster 2, and considers the smallest of these dissimilarities as a linkage criterion.

```{r}
gene <- read_csv("ss.csv")
gene <- scale(gene)
head(gene)
```

```{r}
# Dissimilarity matrix
d <- dist(gene, method = "euclidean")

# Hierarchical clustering using single Linkage
hc1 <- hclust(d, method = "complete" )

# Plot the obtained dendrogram
plot(hc1, cex = 0.6, hang = -1)
```
Ward’s minimum variance method: It minimizes the total within-cluster variance. At each step the pair of clusters with minimum between-cluster distance are merged.

```{r}
# Ward's method
hc5 <- hclust(d, method = "ward.D2" )

# Cut tree into 4 groups
sub_grp <- cutree(hc5, k = 3)

# Number of members in each cluster
table(sub_grp)
```

```{r}
plot(hc5, cex = 0.6)
rect.hclust(hc5, k = 3, border = 2:5)
```
Visualize the result in a scatter plot.
```{r}
fviz_cluster(list(data = gene, cluster = sub_grp))
```

Determining Optimal Clusters
* Elbow Method
```{r}
fviz_nbclust(gene, FUN = hcut, method = "wss")
```
* Average Silhouette Method
```{r}
fviz_nbclust(gene, FUN = hcut, method = "silhouette")
```
* Gap Statistic Method
```{r}
gap_stat <- clusGap(gene, FUN = hcut, nstart = 25, K.max = 10, B = 50)
fviz_gap_stat(gap_stat)
```

## Conclusion