---
title: "Report"
author: "Liucheng Shi, Zhuohui Liang, Ruwen Zhou, Jiying Han"
date: "3/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(caret)
library(parallel)
library(foreach)
library(ggfortify)
library(patchwork)
library(cluster)
library(factoextra) 
library(dendextend) 
library(tidyverse)

knitr::opts_chunk$set(
  fig.height = 6,
  fig.width = 8,
  message = F,
  echo = F,
  warning = F,
  cache = F
)

theme_set(theme_minimal() + theme(legend.position = "bottom", 
                                  title = element_text(hjust = 0.5)))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis",
  digits = 3
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

set.seed(123123)
```

# Introduction


# Method

## Data Preparation

```{r load_data, echo=F}
sngcll_raw = 
  read_csv("ss.csv") 

sngcll = sngcll_raw %>% 
  janitor::clean_names()
```

```{r standardize_data}
# drop gene that 90% is zero 
drop_gene = 
  sngcll %>% 
  summarise(across(everything(),~sum(.x==0)/n()<0.9)) %>% 
  slice(1) %>% 
  unlist() %>% 
  as.vector()

sngcll_dp0 = 
  sngcll[,drop_gene]
```


## PCA

```{r pca}
# PCA with scale
sngcll_pca =
  #predict(preProcess(sngcll_dp0,c("center","scale","pca")),sngcll_dp0)
  prcomp( ~ .,
          data = sngcll_dp0,
          tol = sqrt(.Machine$double.eps),
          center =F,
          scale. = F,
          scale = F)

summary(sngcll_pca)$importance %>% 
  t() %>% 
  .[seq(1,230,8),] %>% 
  as_tibble(rownames = NA) %>% 
  knitr::kable(digits = 2)

sngcll_pca$rotation[,1:4] %>% t()%>% knitr::kable()
```

## Apply EM algorithm

```{r EM_data}
# EM data preparation
sngcll_pca =
  predict(preProcess(
    sngcll_dp0,
    c("pca","center","scale"),
    pcaComp = sum(summary(sngcll_pca)$importance[3, ]<.9)
  ), sngcll_dp0)
```

```{r eval = F}
source("EM.R")

set.seed(123123)
rslt_2 = rerun(1,lapply(10:2, FUN = function(x) gaussian_mixture(sngcll_pca,x,method ="AIC")) %>% do.call(rbind,.)) %>% do.call(rbind,.)
save(rslt_2,file = "EM.Rdata")

set.seed(123123)
t1 = Sys.time()
rslt = rerun(20,lapply(10:2, FUN = function(x) gaussian_mixture(sngcll_pca,x,method ="AIC")) %>% do.call(rbind,.)) %>% do.call(rbind,.)
save(rslt,file = "EM_20.Rdata")
t_run = Sys.time() - t1
```



  The model is assumed to have gaussian conditional distribution in each cluster, with parameters $\mu_k$ and $\Sigma_k$ for k in 1,2,3...K. Each observation have probability $p_{ik}$ to be any of the K's clusters, observation is assigned to the cluster with the highest probability. The model is presented as followed:
  
$$\mathbf x_i\sim
\begin{cases}
N(\boldsymbol \mu_1, \Sigma_1), \mbox{with probability }p_1 \\
N(\boldsymbol \mu_2, \Sigma_2), \mbox{with probability }p_2\\
\quad\quad\vdots\quad\quad,\quad\quad \vdots\\
N(\boldsymbol \mu_k, \Sigma_k), \mbox{with probability }p_k\\
\end{cases}$$


  Where i $\in$ 1,2,3...N, the completed likelihood function is:

$$L(\theta; \mathbf x,\mathbf r) = \prod_{i=1}^n \prod_{j=1}^k [p_j f(\mathbf x_i; \boldsymbol \mu_j, \Sigma_j)] ^{r_{i,j}}$$

  The EM algorithm is designed as:
\begin{tabular}{c c c}
& &describe\\
\hline\\
1& &initialize model with random $p_j,\mu_j,\Sigma_j$ given k\\
2& while& iteration is less than maximum iteration or objective function not converge\\
&\vline& E step: calculate the conditional probability $p(r_j|X,\mu,\Sigma)$\\
&\vline& M step: calculate and update $\mu,\Sigma,p$ and assign clusters\\
&end&
\end{tabular}

  After optimizing, calculate observed Likelihood L as well as the AIC loss function($-2(log(L)-n_p)$). where $n_p=K+GK+G^2K+KN$. As rule of thumb, initial cluster's number is set as 2 to 10\cite{system_review}, and final cluster number is determined by model with lowest AIC.
  
# Result

## Gene-expression signatures

## Analysing Gaussian-Mixture model with Principal Component

```{r plot}
load("EM.Rdata")

ggplot(as_tibble(rslt_2) %>% unnest(c(k, obj)), aes(k, obj)) +
  geom_path()+
  labs(y = "AIC")

sngcll_clu_3 = cbind(sngcll_pca,rslt_2[which(rslt_2[,"k"]==3),"cluster"])

pc2 = ggplot(sngcll_clu_3,aes(PC1,PC2,color = cluster))+geom_jitter(alpha = 0.8)

pc3 = ggplot(sngcll_clu_3,aes(PC1,PC3,color = cluster))+geom_jitter(alpha = 0.8)

pc4 = ggplot(sngcll_clu_3,aes(PC1,PC4,color = cluster))+geom_jitter(alpha = 0.8)

pc2 + pc3+pc4 +plot_spacer()+ plot_layout(nrow = 2,ncol = 2, guides = "collect")
```

## Hierarchical clustering

Hierarchical clustering is an alternative approach to k-means clustering for identifying groups in the dataset. In our project, our purpose is to classify scRNA-seq into different clusters based on their gene expressions and identify potential existence of cell subtypes. We choose agglomerative hierarchical clustering method, which is a bottom-up method. Each object is initially considered as a single-element cluster. At each step of the algorithm, the two clusters that are the most similar are combined into a new bigger cluster. The iteration will not stop until all elements are being classified into one single cluster. This makes the result a tree and can be visualized as a dendrogram.

When measuring the dissimilarity between each pair of observations distance, Euclidean distance has been used. However, when measuring the dissimilarity between two clusters of observations, we have applied different methods.

* Minimum or single linkage clustering: It computes all pairwise dissimilarities between the elements in cluster 1 and the elements in cluster 2, and considers the smallest of these dissimilarities as a linkage criterion.

```{r}
gene <- read_csv("ss.csv")
gene <- scale(gene)
head(gene)
```

```{r}
# Dissimilarity matrix
d <- dist(gene, method = "euclidean")

# Hierarchical clustering using single Linkage
hc1 <- hclust(d, method = "complete" )

# Plot the obtained dendrogram
plot(hc1, cex = 0.6, hang = -1)
```
Ward’s minimum variance method: It minimizes the total within-cluster variance. At each step the pair of clusters with minimum between-cluster distance are merged.

```{r}
# Ward's method
hc5 <- hclust(d, method = "ward.D2" )

# Cut tree into 4 groups
sub_grp <- cutree(hc5, k = 3)

# Number of members in each cluster
table(sub_grp)
```

```{r}
plot(hc5, cex = 0.6)
rect.hclust(hc5, k = 3, border = 2:5)
```
Visualize the result in a scatter plot.
```{r}
fviz_cluster(list(data = gene, cluster = sub_grp))
```

Determining Optimal Clusters
* Elbow Method
```{r}
fviz_nbclust(gene, FUN = hcut, method = "wss")
```
* Average Silhouette Method
```{r}
fviz_nbclust(gene, FUN = hcut, method = "silhouette")
```
* Gap Statistic Method
```{r}
gap_stat <- clusGap(gene, FUN = hcut, nstart = 25, K.max = 10, B = 50)
fviz_gap_stat(gap_stat)
```

## Conclusion